A Higher-Order Approach to Explicit Contexts in LF

### 2015-03-08 paper/paper.tex
corrected operational semantics
stub out admissibility lemmas
typo
[paper] Proved admissibility of function application rule

\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newdef{definition}[theorem]{Definition}
\newdef{remark}[theorem]{Remark}
\def\permission{}

@@

\begin{gather*}
  \infer{
    \eval{M}{N}:\type
  }{
    M:\sortexp &
    N:\sortval
  }\\[6pt]
  %
  \infer[\evalrule\uparrow]{
    \eval{\thunk{M}}{M}
  }{
  }
  \quad
  \infer[\evalrule\ap]{
    \eval{\ap(\thunk{\lam[x]E(x)};M)}{N}
  }{
    \eval{E(M)}{N}
  }\\[6pt]
  %
  \infer[\evalrule\fst]{
    \eval{\fst(\thunk{\pair(M;N)})}{M'}
  }{
    \eval{M}{M'}
  }
  \quad
  \infer[\evalrule\snd]{
    \eval{\fst(\thunk{\pair(M;N)})}{N'}
  }{
    \eval{N}{N'}
  }
\end{gather*}

@@

Let us recall Martin-L\"of's informal rules and we will show how they are
encoded in the logical framework. First, the cartesian product of a family of
sets is defined by the following rules in \cite{Martin-Lof-1979}:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Pi}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(\lambda x)b}{(\mathsf{\Pi}x\in A)B}
  }{
    \mem{b}{B}\ (\mem{x}{A})
  }\quad
  \infer{
    \eqmem{(\lambda x)b}{(\lambda x)d}{(\mathsf{\Pi}x\in A)B}
  }{
    \eqmem{b}{d}{B}\ (x\in A)
  }
\end{gather*}
%
These rules are adequately encoded in our system as follows:
\begin{gather*}
  \infer{
    \canset{\typi(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\lam(E)}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
  }\quad
  \infer{
    \eqcanmem{\lam(E)}{\lam(E')}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\eqcanmem{\tacitbind{E}}{\tacitbind{E'}}{\tacitbind{B}}}
  }
\end{gather*}

@@

>>>>

\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{case}{Case}
\newdef{definition}[theorem]{Definition}
\newdef{remark}[theorem]{Remark}
\def\permission{}

@@

\begin{gather*}
  \infer{
    \eval{M}{N}:\type
  }{
    M:\sortexp &
    N:\sortval
  }\\[6pt]
  %
  \infer[\evalrule\uparrow]{
    \eval{\thunk{M}}{M}
  }{
  }
  \quad
  \infer[\evalrule\ap]{
    \eval{\ap(L;M)}{N}
  }{
    \eval{L}{\lam[x]E(x)} &
    \eval{E(M)}{N}
  }\\[6pt]
  %
  \infer[\evalrule\fst]{
    \eval{\fst(M)}{M_1'}
  }{
    \eval{M}{\pair(M_1;M_2)} &
    \eval{M_1}{M_1'}
  }
  \quad
  \infer[\evalrule\snd]{
    \eval{\snd(M)}{M_2'}
  }{
    \eval{M}{\pair(M_1;M_2)} &
    \eval{M_2}{M_2'}
  }
\end{gather*}

@@

Let us recall Martin-L\"of's informal rules and we will show how they are
encoded in the logical framework. First, the cartesian product of a family of
sets is defined by the following rules in \cite{Martin-Lof-1979}:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Pi}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(\lambda x)b}{(\mathsf{\Pi}x\in A)B}
  }{
    \mem{b}{B}\ (\mem{x}{A})
  }\quad
  \infer{
    \eqmem{(\lambda x)b}{(\lambda x)d}{(\mathsf{\Pi}x\in A)B}
  }{
    \eqmem{b}{d}{B}\ (x\in A)
  }
\end{gather*}
%
These rules are adequately encoded in our system as follows:
\begin{gather*}
  \infer{
    \canset{\typi(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\lam E}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
  }\quad
  \infer{
    \eqcanmem{\lam E}{\lam E'}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\eqcanmem{\tacitbind{E}}{\tacitbind{E'}}{\tacitbind{B}}}
  }
\end{gather*}

@@

\subsection{Admissible Rules}

\theorem{The standard application rule for the cartesian product of a family of
sets is admissible:
\[
  \infer[\typi_E]{
    \mem{\ap(M;N)}{B(N)}
  }{
    \begin{array}{c}
      \mathcal{D}\\
      \mem{M}{\thunk{\typi(A;B)}}\\
    \end{array} &
    \begin{array}{c}
      \mathcal{E}\\
      \mem{N}{A}
    \end{array}
  }
\]
}
\begin{proof}
  By induction on the derivation $\mathcal{D}$, which must have the following canonical form:
  \[
     \infer[\mathcal{D}]{
       \mem{M}{\thunk{\typi(A;B)}}
     }{
       \begin{array}{c}
         \mathcal{D}' \\
         \eval{M}{\lam E}
       \end{array} &
       \infer{
         \canmem{\lam E}{\typi(A;B)}
       }{
         \infer{
           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
         }{
           \infer{
             \lfpi{x,x'}{
               \weaken{\cdot}{x}{x'}
               \to\sequent{\cdot}{\qmem{x'}{A}}
               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
             }
           }{
             \begin{array}{c}
               \mathcal{D}''\\
               \lfpi{x}{\mem{x}{A} \to \mem{E(x)}{B(x)}}
             \end{array}
           } &
           \dots
         }
       }
     }
  \]

  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
  \[
    \infer{
      \mem{E(N)}{B(N)}
    }{
      \begin{array}{c}
        \mathcal{F}\\
        \eval{E(N)}{X}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
  Whence we may construct the goal:
  \[
    \infer{
      \mem{\ap(M;N)}{B(N)}
    }{
      \infer{
        \eval{\ap(M;N)}{X}
      }{
        \infer[\mathcal{D}']{\eval{M}{\lam E}}{} &
        \infer[\mathcal{F}]{\eval{E(N)}{X}}{}
      }&
      \infer[\mathcal{F}']{\eval{B(N)}{C}}{}&
      \infer[\mathcal{F}'']{\canmem{X}{C}}{}&
    }
  \]
\end{proof}

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
  }
\]
}
\begin{proof}
\end{proof}
